### Matrices can have both column and row names.

```{r}
m <- matrix(1:4, nrow = 2, ncol = 2)
m
dimnames(m) <- list(c("a", "b"), c("c", "d")) 
m
```

```{r}
# the comma separates row and columns
# without any indexing before of after the comma, the outpout would be:
# ALL rows and ALL columns
m[,]
```

```{r}
m[1, ] # First Row
```

```{r}
m[, 1] # First Column
```

```{r}
m["a","c"] # Row a, Column c
```

### Column names and row names can be set separately using the `colnames()` and `rownames()` functions.

```{r}
colnames(m) <- c("h", "f")
rownames(m) <- c("x", "z")
m
```

## Data Frames

### Most often Data Frames are created while importing your data files using function like read.csv (base-r) or read_csv (tibble).

### But Data Frames can also be created manually in different ways:

```{r}
  x <- data.frame(foo = 1:4, bar = c(T, T, F, F)) 
  x

```

```{r}
# Creating individual lists
# Remember: each vector must be a unique data-type
name <- c("Alice", "Bob", "Charlie")
age <- c(25, 30, 22)
score <- c(88, 75, 92)

# Combining lists to create a DataFrame
data <- data.frame(Name = name, Age = age, Score = score)

# Displaying the DataFrame
print(data)

```

#### Create a Data Frame by converting it from another similar structure e.g. Matrix:

Often done to process some numerical operations first (matrix):

```{r}
m <- matrix(1:6, nrow = 2, ncol = 3) 
m
```

```{r}
my_df <- as.data.frame(m)

my_df
```

```{r}
colnames(my_df) <- c("col1","col2","col3")
my_df
```

## Simulate a gene expression Data-Frame

```{r}
# Set a seed for reproducibility (optional)
set.seed(42)

# Define the number of genes and samples
num_genes <- 100  # Number of genes
num_samples <- 10  # Number of samples

# Create the dataframe with random expression values
gene_names <- paste0("Gene_", 1:num_genes)
expression_data <- data.frame(
  Gene = gene_names,
  sample = matrix(rnorm(num_genes * num_samples, mean = 50, sd = 10), nrow = num_genes)
)

# View the dataframe
print(expression_data)

```

```{r}
nrow(expression_data)
ncol(expression_data)
```

```{r}
rownames(expression_data) <- expression_data$Gene

expression_data$Gene <- NULL

head(expression_data)
```

```{r}
my_mtrx <- as.matrix(expression_data)
head(my_mtrx)
```

```{r}
expression_data
```

```{r}
library(pheatmap)

pheatmap(expression_data[, 2:10])
```

```{r}
head(expression_data) # head is to show the beginning of a dataframe

tail(expression_data) # shows the ending of a dataframe
```

There are various functions that can give you an overview of the data that is in a data frame, which may or may not be useful depending on what type of data you have:

| Function         | Returns                                    |
|:-----------------|:-------------------------------------------|
| `colnames(df)`   | The column names of a data frame           |
| `rownames(df)`   | The row names of a data frame              |
| `ncol(df)`       | Number of columns                          |
| `nrow(df)`       | Number of rows                             |
| `describeBy(df)` | Descriptive statistics                     |
| `summary(df)`    | Results of various model fitting functions |

```{r}
expression_data[1:5 , "sample.1"]

```

```{r}
summary(expression_data[,"sample.1"])
```

```{r}
# an column can be easily removed:
# remove the sample.1
expression_data$sample.1 <- NULL

head(expression_data)
```

```{r}

rownames(expression_data) <- expression_data$Gene

expression_data$Gene <- NULL

pheatmap(expression_data)
```

```{r}


# Download the data from GitHub and save it as 'data.csv' in your working directory
download.file("https://raw.githubusercontent.com/oHunewald/NI2_Training/main/part2/expr_2.csv", "expr_2.csv")

# Read the downloaded CSV file into a dataframe
data <- read.csv("expr_2.csv", sep = ";", dec=",")

# View the dataframe
head(data)

# another way how to peek into a large dataset with many rows and columns is by sub-setting to 
# less rows and columns, here for example, we limit the output to 5 rows and 5 columns
data[1:5,1:5]
```

```{r}

# check the dimension of the dataset (output is number of rows and number of columns)
dim(data)
```

```{r}
# using a heatmap function to visualize the expression data
pheatmap(data)
```

As the ranges of marker intensities can vary substantially, we apply another transformation that scales expression of all markers to values between 0 and 1 using low (e.g. 1%) and high (e.g. 99%) percentiles as the boundary.

```{r}
# if needed install the library first
# install.packages("matrixStats")

library(matrixStats)
expr <- as.matrix(data)
rng <- colQuantiles(expr, probs = c(0.01, 0.99))
expr01 <- t((t(expr) - rng[, 1]) / (rng[, 2] - rng[, 1]))
expr01[expr01 < 0] <- 0
expr01[expr01 > 1] <- 1
```

```{r}
pheatmap(expr01)
